---
title: "Introduction to ssstats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ssstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

The `ssstats` package provides a consistent set of functions for applied statistics courses at the University of West Florida. Every function follows the same conventions: pass a data frame, specify variable names without quotes, and receive formatted output that walks through hypotheses, test statistics, and conclusions.

To follow along, load the required packages:

```{r}
library(tidyverse)
library(ssstats)
library(palmerpenguins)
```

All examples in this vignette use the `penguins` dataset from `palmerpenguins`, which contains measurements for 344 penguins across three species.

```{r}
penguins <- penguins
head(penguins)
```

---

## Descriptive Statistics

### Means and Medians

`mean_median()` produces a summary of mean (SD) and median (IQR) for one or more numeric variables.

```{r}
mean_median(penguins, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)
```

### Frequency Tables

`n_pct()` produces a one- or two-way frequency table with counts and column percentages.

```{r}
# One-way table
n_pct(penguins, species)
```

```{r}
# Two-way table
n_pct(penguins, species, island)
```

### Descriptive Statistics Table

`table_one()` creates a publication-style descriptive statistics table. Continuous variables are summarized as mean (SD) or median (IQR); categorical variables show counts and column percentages. Use `split_var` to stratify by a grouping variable.

```{r}
table_one(
  penguins,
  cont_vars = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
  cat_vars  = c(species, island),
  split_var = sex
)
```

---

## One-Sample Procedures

### Checking Normality

Before running a one-sample t-test, use `one_qq_plot()` to assess whether the data are approximately normally distributed.

```{r}
one_qq_plot(penguins, bill_length_mm)
```

### One-Sample Mean (t-test)

`one_mean()` tests whether the population mean equals a hypothesized value using a one-sample t-test.

```{r}
one_mean(penguins, bill_length_mm, mu = 43)
```

### One-Sample Proportion (z-test)

`one_proportion()` tests a single proportion using a z-test with a Wald confidence interval. Specify the `event` value that counts as a success.

```{r}
one_proportion(penguins, sex, event = "male", p = 0.5)
```

### One-Sample Variance (chi-squared test)

`one_variance()` tests whether the population variance equals a hypothesized value using a chi-squared test.

```{r}
one_variance(penguins, bill_length_mm, sigma2 = 25)
```

---

## Two-Sample Independent Procedures

The examples in this section compare Adelie and Chinstrap penguins, filtering out the third species.

```{r}
ac_penguins <- penguins |> filter(species != "Gentoo")
```

### Checking Normality by Group

`independent_qq()` produces QQ plots and histograms for each group, useful for assessing the normality assumption before running an independent-samples t-test.

```{r fig.height=5}
independent_qq(ac_penguins, bill_length_mm, species)
```

### Two-Sample Means (Welch's t-test)

`independent_means()` compares means from two independent groups using Welch's t-test (unequal variances assumed). Use `first` to control which group is subtracted from.

```{r}
independent_means(ac_penguins, bill_length_mm, species, first = "Chinstrap")
```

### Two-Sample Variances (F-test)

`two_variances()` tests the equality of variances between two groups.

```{r}
two_variances(ac_penguins, bill_length_mm, species)
```

### Two-Sample Medians (Wilcoxon Rank Sum Test)

`independent_medians()` provides a non-parametric alternative based on the Wilcoxon rank sum test, reporting the Hodges-Lehmann estimate of the location shift.

```{r}
independent_medians(ac_penguins, bill_length_mm, species, first = "Chinstrap")
```

### Two-Sample Proportions (z-test)

`independent_proportions()` compares proportions between two groups. Here we test whether the proportion of female penguins differs between Adelie and Chinstrap.

```{r}
ac_penguins |>
  filter(!is.na(sex)) |>
  independent_proportions(sex, species, event = "female", first = "Chinstrap")
```

---

## Paired (Dependent) Procedures

### Checking Normality of Differences

`dependent_qq()` produces a QQ plot and histogram of the paired differences to assess normality.

```{r fig.height=5}
dependent_qq(penguins, bill_length_mm, bill_depth_mm)
```

### Paired Means (Paired t-test)

`dependent_means()` tests the mean of paired differences using a paired t-test.

```{r}
dependent_means(penguins, bill_length_mm, bill_depth_mm)
```

### Paired Medians (Wilcoxon Signed-Rank Test)

`dependent_medians()` provides a non-parametric alternative using the Wilcoxon signed-rank test.

```{r}
dependent_medians(penguins, bill_length_mm, bill_depth_mm)
```

### Paired Proportions (McNemar's Test)

`dependent_proportions()` tests paired binary outcomes using McNemar's test. It automatically uses the normal approximation when there are 25 or more discordant pairs, and an exact binomial test otherwise.

The following example uses a small before/after survey dataset:

```{r}
survey <- tibble(
  pre  = c("yes","yes","no","yes","no","no","yes","no","yes","yes",
           "no","yes","no","no","yes","yes","no","yes","no","no"),
  post = c("yes","no","no","yes","yes","no","yes","yes","yes","no",
           "yes","yes","no","no","yes","yes","yes","no","no","yes")
)

dependent_proportions(survey, pre, post, event = "yes")
```

---

## Multiple-Group Procedures

### Checking ANOVA Assumptions

Before running an ANOVA, use `anova_check()` to examine residuals for normality (QQ plots faceted by group) and equal variance.

```{r fig.height=5}
anova_check(penguins, bill_length_mm, species)
```

### One-Way ANOVA

`anovas()` fits and summarizes a one-way ANOVA when only factor `A` is provided.

```{r}
anovas(penguins, bill_length_mm, species)
```

### Post-Hoc Comparisons

After a significant ANOVA, use `posthoc_test()` to identify which group pairs differ. By default it applies Tukey's HSD adjustment; set `adjust = FALSE` for Fisher's LSD.

```{r}
posthoc_test(penguins, bill_length_mm, A = species)
```

### Simultaneous Confidence Intervals

`simultaneous_ci()` constructs Bonferroni-corrected confidence intervals for a parameter (mean, median, variance, SD, or proportion) across all groups.

```{r}
simultaneous_ci(penguins, bill_length_mm, species)
```

### Two-Way ANOVA

Pass a second factor `B` to `anovas()` to fit a two-way ANOVA. Set `interaction = TRUE` (default) to include the interaction term.

```{r}
penguins |>
  filter(!is.na(sex)) |>
  anovas(bill_length_mm, species, sex, interaction = TRUE)
```

Use `profile_plot()` to visualize potential interactions before (or alongside) a two-way ANOVA:

```{r}
penguins |>
  filter(!is.na(sex)) |>
  profile_plot(bill_length_mm, xaxis = species, lines = sex)
```

### Kruskal-Wallis Test

`kruskal_wallis()` provides a non-parametric alternative to one-way ANOVA when normality is not tenable.

```{r}
kruskal_wallis(penguins, bill_length_mm, species)
```

For post-hoc comparisons after Kruskal-Wallis, set `nonparametric = TRUE` in `posthoc_test()`:

```{r}
posthoc_test(penguins, bill_length_mm, A = species, nonparametric = TRUE)
```

### Brown-Forsythe-Levene Test

`multiple_variances()` tests the equality of variances across three or more groups using the median-based Brown-Forsythe-Levene test.

```{r}
multiple_variances(penguins, bill_length_mm, species)
```

---

## Categorical Data Analysis

### Chi-Square Goodness-of-Fit

`goodness_of_fit()` tests whether observed category frequencies match a specified distribution. If `expected` is `NULL`, equal proportions are assumed.

```{r}
goodness_of_fit(penguins, species)
```

To test against a non-uniform distribution, supply a named vector of expected proportions:

```{r}
goodness_of_fit(
  penguins,
  species,
  expected = c(Adelie = 0.50, Chinstrap = 0.25, Gentoo = 0.25)
)
```

### Chi-Square Test of Independence

`independence()` tests whether two categorical variables are associated.

```{r}
independence(penguins, species, island)
```

---

## Correlation

### Checking Bivariate Normality

`correlation_qq()` produces side-by-side QQ plots for two variables to help assess whether the normality assumption for Pearson's correlation is plausible.

```{r}
correlation_qq(penguins, bill_length_mm, body_mass_g)
```

### Pearson and Spearman Correlation

`correlation()` computes Pearson (default) or Spearman correlation. Provide both `x` and `y` for a single pair, or omit them for all pairwise correlations.

```{r}
# Single pair — Pearson
correlation(penguins, bill_length_mm, body_mass_g)
```

```{r}
# Single pair — Spearman
correlation(penguins, bill_length_mm, body_mass_g, method = "spearman")
```

```{r}
# All numeric pairs
penguins |>
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  correlation()
```

---

## Regression

### Is the Line Significant? (Likelihood Ratio Test)

`significant_line()` tests the overall significance of a regression model via a likelihood ratio test comparing the fitted model to an intercept-only null model.

```{r}
significant_line(penguins, body_mass_g, bill_length_mm)
```

### Linear Regression Table

`linear_regression()` fits a linear model and returns a formatted table of coefficients, 95% confidence intervals, and p-values.

```{r}
linear_regression(penguins, body_mass_g, bill_length_mm)
```

For multiple predictors, pass an expression to `function_of`:

```{r}
linear_regression(penguins, body_mass_g, bill_length_mm + species)
```

### Adjusted R²

`r_squared()` returns the adjusted R² for a fitted model or from raw data.

```{r}
r_squared(data = penguins, outcome = body_mass_g, function_of = bill_length_mm + species)
```

### Comparing Models Side by Side

`glm_table()` accepts multiple fitted model objects and produces a single comparison table.

```{r}
m1 <- lm(body_mass_g ~ bill_length_mm, data = penguins)
m2 <- lm(body_mass_g ~ bill_length_mm + species, data = penguins)

glm_table(m1, m2, model_names = c("Simple", "With Species"))
```

### Checking Regression Assumptions

`reg_check()` produces three diagnostic plots: a QQ plot of residuals, a histogram of residuals, and a residuals vs. fitted values plot. It accepts either a fitted model object or raw data with column names.

```{r fig.height=5}
reg_check(m2)
```

`normality_check()` and `variance_check()` isolate the individual components if you prefer to view them separately:

```{r}
normality_check(m2)
```

```{r}
variance_check(m2)
```

### Cook's Distance

`cooks()` plots Cook's distance for each observation and flags those that exceed a threshold. The default `"convention"` threshold uses $4/n$ as a low threshold and 1 as a high threshold.

```{r}
cooks(m2)
```

### Outlier Detection

`outlier_count()` tabulates how many observations have absolute standardized residuals above 2.5 (suspected outliers).

```{r}
outlier_count(m2)
```

`outlier_graph()` produces a scatter plot colored by outlier status:

```{r}
outlier_graph(penguins, m1, bill_length_mm, body_mass_g)
```

### Logistic Regression

`logistic_regression()` fits a logistic regression model and returns a table of log-odds coefficients, confidence intervals, and p-values. The outcome variable should be coded as 0/1 (or a factor with two levels).

```{r}
penguins |>
  filter(!is.na(sex)) |>
  mutate(male = as.integer(sex == "male")) |>
  logistic_regression(male, bill_length_mm + bill_depth_mm)
```
